---
title: "Exercises for Spatial Point Processes"
subtitle: "Spatial Epidemiologi Autumn 2021"
author: "Alexander, Victor, Mikel"
date: "`r format(Sys.time(), '%d %B, %Y')`"
params: 
  show_code: FALSE
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    theme: readable
    highlight: textmate
    number_sections: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, warning = F)
rm(list = ls())
setwd("/home/ajo/gitRepos/spatial/spp/Assignment")
library(spatstat)
```

# Task 1

## Data import and cleaning

```{r data-import}
nucli.23=read.table(file="nucli23.txt",header=T,sep="\t")
head(nucli.23)
summary(nucli.23)

poly.23=read.table(file=paste("poly23.txt"),header=T,sep="\t")
summary(poly.23)

# Change reference of X and Y (to 0). 
min.X=min(nucli.23$X)
min.Y=min(nucli.23$Y)
nucli.23$X=nucli.23$X-min.X
nucli.23$Y=nucli.23$Y-min.Y
summary(nucli.23)

poly.23$X=poly.23$X-min.X
poly.23$Y=poly.23$Y-min.Y
summary(poly.23)
```

## 1.1) Build a ppp object using the “poly23” data as a window. 

```{r}
pol.illa<-list(x=poly.23$X,y=poly.23$Y)
n23=ppp(nucli.23$X,nucli.23$Y,poly=pol.illa)
plot(n23,main="nucli23")
axis(1,at=c(seq(0,75,by=25)),pos=c(-10,0))
axis(2,at=c(seq(0,100,by=25)),pos=c(-10,-0))
```

## 1.2) Briefly describe the point pattern process. 

```{r}
summary(n23)
n23.den<-density(n23,dimyx=c(256,256))
plot(n23.den)
plot(density(n23, dimyx=c(256,256), sigma=1))
plot(density(n23, dimyx=c(256,256), sigma=5))
plot(density(n23, dimyx=c(256,256), sigma=7.5))
axis(1,at=c(seq(0,75,by=25)),pos=c(-10,0))
axis(2,at=c(seq(0,100,by=25)),pos=c(-10,-0))
plot(density(n23, dimyx=c(256,256), sigma=10))
plot(density(n23, dimyx=c(256,256), sigma=20))
plot(density(n23, dimyx=c(256,256), sigma=50))
```

We can see that sigma = 20 and sigma = 50 gives too much smoothing. Similarly, we can see that sigma = 1 gives to little smoothing. The density with sigma between 5 and 10 looks more informative. 

Additionally, we can see that the process is not homogeneous. There are some regions that clearly have a larger intensity of points. The regions around $(x,y) = (60, 25)$ and $(x,y) = (30,60)$ are two examples, where the first one has the largest intensity. 

## 1.3) Rebuild the ppp object using the “time to nest” (data_pos variable) as marks.

```{r}
(n23T=ppp(nucli.23$X,nucli.23$Y,poly=pol.illa,marks=nucli.23$data_pos))
```

## 1.4) Briefly describe the marked point process. 

```{r}
summary(n23T)
plot(n23T,main="nucli23",markscale=0.2,leg.side="right")
axis(1,at=c(seq(0,75,by=25)),pos=c(-10,0))
axis(2,at=c(seq(0,100,by=25)),pos=c(-10,-0))
# Hard to interpret, categorize time into categories for every 4 days instead. 

DPOScat=cut(nucli.23$data_pos,breaks=c(9, 12, 14,16, 18, 20, 23),labels=c("10-12","12-14","14-16","16-18", "18-20", "20-22"))
table(DPOScat) # Are no points 12-14. Thus, merge 10-12 and 12-14 into 10-14.
DPOScat=cut(nucli.23$data_pos,breaks=c(9, 14,16, 18, 20, 22),labels=c("10-14","14-16","16-18", "18-20", "20-22"))
table(DPOScat) # Are no points 12-14. Thus, merge 10-12 and 12-14 into 10-14.
(n23Tcat=ppp(nucli.23$X,nucli.23$Y,poly=pol.illa,marks=DPOScat))
summary(n23Tcat)
plot(split(n23Tcat))

plot(n23Tcat,main="nucli23",cex=0.75,cols=c("purple", "green", "blue","red", "black"),
     chars=rep(16,4),leg.side="right")
axis(1,at=c(seq(0,75,by=25)),pos=c(-10,0))
axis(2,at=c(seq(0,100,by=25)),pos=c(-10,-0))

plot(density(split(n23Tcat), sigma = 7.5), ribbon = F)

nucli.23.df=data.frame(nucli.23,DPOScat)
n23new=split(nucli.23.df,DPOScat)
for (i in 1:5){
  temp.ppp=ppp(n23new[[i]]$X,n23new[[i]]$Y,poly=pol.illa)
  plot(temp.ppp,pch=i,main=c("Group",i))
  axis(1,at=c(seq(0,75,by=25)),pos=c(-10,0))
  axis(2,at=c(seq(0,100,by=25)),pos=c(-10,-0))
  par(ask=T)
}
par(ask=F)
```

It is hard to see a clear pattern in how the point process evolves with time, but we can make some remarks: 
It looks like the area we noted as high intensity at first ((x,y) = (60, 25)) stays high intensity through all times that were recorded. One can think that it never gets saturated with nests, i.e. it is a good area for nesting, with room,throughout all recorded time. The other area we noted as relatively high intensity ((x,y) = (30,60)) is most "popular" for nesting between weeks 16 and 20. There is also an area around (x,y) = (10,90) (in the north-west) that has some nests in all weeks.

# Task 2

## 2.1) Give a point estimate and a 95% confidence interval of the intensity, assuming a homogeneous Poisson process. 

When assuming a homogeneous Poisson process, a point estimate of the intensity is simply the average intensity in the window. This means that the point estimate can be read from the summary of the ppp-object, 

```{r}
summary(n23)
```

i.e. the point estimation is `r summary(n23)$intensity`. The confidence interval can be found from the results shown below **I think!!??**

```{r}
model1=ppm(n23, ~1)
model1
```


## 2.2) Assess the Completely Spatial Randomness hypothesis. 

The Completely Spatial Randomness (CSR) hypothesis can be assessed with a chi-squared test based on quadrat counts or a Kolmogorov-Smirnov test. First, let us perform the chi-squared test

```{r}
# Tried to set own brakes, but did not get it to work properly. 
#M <- quadrat.test(n23, xbreaks = c(0, 20, 40, 60), ybreaks = c(20, 40, 60, 80, 100))
M <- quadrat.test(n23, nx = 2, ny = 3)
M
plot(n23,main="nucli23")
plot(M, add = TRUE)
axis(1,at=c(seq(0,75,by=25)),pos=c(-10,0))
axis(2,at=c(seq(0,100,by=25)),pos=c(-10,-0))
```

Since the test above has a very large $p$-value, we cannot conclude that this is a completely random spatial process, since we cannot reject the null hypothesis of CSR. **Not sure if this makes sense right now, think more**

http://rstudio-pubs-static.s3.amazonaws.com/5292_2b2fae3795a144b2a4b486fd2fc6fc57.html

Next, the Kolmogorov-Smirnov test is performed. 

```{r}
KS=cdf.test(n23,covariate="x")
KS
plot(KS)

KS=cdf.test(n23,covariate="y")
KS
plot(KS)
```

The $p$-values from this test are also quite large, which means that the Kolmogorov-Smirnov test does not give enough evidence against the null hypothesis of CSR.

Thus, both these two tests lead to the same conclusion; we cannot reject the null-hypothesis of CSR, i.e. we have no evidence of this process not being CSR. 
  
## 2.3) Fit an inhomogeneous Poisson model to data. 

```{r}
# fit an homogeneous poisson model
model1=ppm(n23, ~1)
model1
plot(model1)

# inhomogeneous Poisson model
model2=ppm(n23, ~x + y)
model2
plot(model2, se=F)
```

**Look at some residuals or something as well?**

# Task 3
